---
title: "Data Preprocessing of Population and Life expectancy based on Gender"
author: "Deepthi Suresh"
date: "13 Oct 2023"
output:
  pdf_document: 
  html_notebook: default
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

## **Setup**

```{r}

# Load the necessary packages required to reproduce the report. For example:

library(readr)
library(dplyr)
library(tidyr)
library(magrittr)
library(openxlsx)
library(readxl)
library(Hmisc)
library(editrules)
library (MVN)
library(forecast)
library(kableExtra)
library(outliers)

```

<br> <br>

## **Executive Summary**

The report is based on four datasets - Population and Life Expectancy
(LE) for both males and females. As a part of the pre-processing the
first step was to read in the data into R and then make adjustments to
the data especially removing the 'world' observation to avoid skewness.
The data was then tidied by converting the year into a variable with its
own column. Then the male and female datasets of population and LE was
combined using rbind. The structure of the dataset was then looked at
and necessary conversion of data type was done. Then, the two data was
ready to be merged using join function to form the merged dataset. Once
the dataset was ready, a new derived variable called relative Life
expectancy was created, which was based on the average global life
expectancy. The final data was then scanned for missing values and
imputation or deletion was performed based on the variable to ensure
there was no missing values. Then the dataset was checked for any
outliers and the numeric variable had some outliers which were dealt by
capping or Winsorising method by using a special function. Once the
missing values are outliers are handled, the data was now ready for
transformation. Several data transformation techniques were performed to
identify the best one to convert the data into a normal distribution.
Scaling and Centering was also performed to enhance the comparability

## **Data**

The first dataset considered here is of the population of 265 countries
in the world and the total world population divided into two datasets
based on the gender under the names pop_female (female population across
countries) and pop_male (male population across countries). The second
dataset has the details of the life expectancy at birth of males and
females across different countries, divided into two datsets called
le_female(Life expectancy at birth, female (years)) and le_male (Life
expectancy at birth, male (years)).

The variables in all 4 datasets are given below:

Country Name - Name of the countries that are assessed

Country Code - abbreviation given to each country

Indicator Name - The variable for which the data is collected.

Population or life expectancy of the female or males

Indicator Code - code for identifying the dataset. most likely used
internally

Years from 1960 - 2022- the data is assessed for the years 1960-2022.

The data is sourced from the world bank website with the below URL:

Population, female
-<https://data.worldbank.org/indicator/SP.POP.TOTL.FE.IN?locations=AU>

Population, male -
<https://data.worldbank.org/indicator/SP.POP.TOTL.MA.IN?locations=AU>

Life expectancy, female
-<https://data.worldbank.org/indicator/SP.DYN.LE00.FE.IN?locations=AU>

Life expectancy,
male-<https://data.worldbank.org/indicator/SP.DYN.LE00.MA.IN?locations=AU>

Since the dataset contains the total world as one of the observation, I
have removed it using the filter function as the main focus of the study
is to understand the country-wise variable distribution ( it avoids the
skewness of data).

Head function is used to make sure the data is correctly loaded.

```{r}
# reading in population data for males and females
pop_female <- read_excel("Population_female.xls",skip = 3)
pop_male <- read_excel("Population male.xls",skip = 3)
# reading in Life expectancy data for males and females
le_male <- read_excel("Life expectancy male.xls",skip = 3)
le_female <- read_excel("Life expectancy female.xls",skip = 3)

#removing the world observation to avoid skewness of data
pop_female <- pop_female %>% filter(`Country Name` != "World") 

pop_male <- pop_male %>% filter(`Country Name` != "World") 

le_female <- le_female %>% filter(`Country Name` != "World") 

le_male <- le_male %>% filter(`Country Name` != "World") 

# checking if the data is loaded correctly
head(pop_female, n=3)
head(pop_male, n=3)
head(le_female,n=3)
head(le_male,n=3)

```

## **Tidy & Manipulate Data I**

```{r}
#tidying population data based on year
PFemale_tidy <- pop_female %>% pivot_longer(names_to = "Year", values_to = "Population", cols = 5:67)
head(PFemale_tidy,n=3)
PMale_tidy <- pop_male %>% pivot_longer(names_to = "Year", values_to = "Population", cols = 5:67)
head(PMale_tidy,n=3)
#combining male and female population datset into 1
Population <- rbind(PMale_tidy,PFemale_tidy)
head(Population,n=3)
#tidying life expectancy data based on year
le_female_tidy <- le_female %>% pivot_longer(names_to = "Year", values_to = "Life expectancy at birth", cols = 5:67)
head(le_female_tidy,n=3)
le_male_tidy <- le_male %>% pivot_longer(names_to = "Year", values_to = "Life expectancy at birth", cols = 5:67)
head(le_male_tidy,n=3)
#combining male and female population and LE dataset into 1
Population <- rbind(PMale_tidy,PFemale_tidy)
LE <- rbind(le_female_tidy,le_male_tidy)
head(LE,n=3)
```

All the datasets in this study has all the years from 1960-2022 in
separate columns which indicated that the data is untidy, as it doesn't
meet the principle of 'every variable must have its own column'. In
order to tidy the dataset, pivot_longer() function has been used on both
population and LE of females and males separately. The head() function
shows that the dataset is now tidy and ready to be combined.

The Population and LE data is created by combining the female and male
population and life expectancy data respectively using the rbind()
function. the head() function shows that the process of combining is
successful. LE and Population now has 33390 observations with 6
variables each

<br> <br> 

## **Understand**

```{r}
#checking and updating the structure of Population dataset
str(Population)
#converting country code into factor variable and checking the levels
Population$`Country Code` <- as.factor(Population$`Country Code`)
#converting Indicator name into factor variable and checking the levels and converting the labels to Male and "Female"
unique(Population$`Indicator Name`)
Population$`Indicator Name`<- factor(Population$`Indicator Name`,levels= c("Population, male","Population, female"),
              labels=c("Male","Female"),ordered = TRUE)
levels(Population$`Indicator Name`)
class(Population$`Indicator Name`)
#converting country Name into factor variable and checking the levels
Population$`Country Name` <- as.factor(Population$`Country Name`)
class(Population$`Country Name`)
#checking the updated structure of the dataset
str(Population)
#checking and updating the structure of Life Expectancy dataset
str(LE)
#converting country code into factor variable and checking the levels
LE$`Country Code` <- as.factor(LE$`Country Code`)
#converting Indicator name into factor variable and checking the levels and converting the labels to Male and "Female"
unique(LE$`Indicator Name`)
LE$`Indicator Name`<- factor(LE$`Indicator Name`,levels= c("Life expectancy at birth, male (years)","Life expectancy at birth, female (years)"),
                      labels=c("Male","Female"),ordered = TRUE)
levels(LE$`Indicator Name`)
class(LE$`Indicator Name`)
#converting country Name into factor variable and checking the levels
LE$`Country Name` <- as.factor(LE$`Country Name`)
class(LE$`Country Name`)
#checking the updated structure of the dataset
str(LE)
#Merging the population and LE dataset together to form Merged dataset
Merged <- inner_join(Population,LE, by = c ("Country Code", "Indicator Name", "Year", "Country Name"))
# Removing the indicator code variable as it is irrelevant to the study and create a cleaner datset
Merged <- Merged %>% select(-`Indicator Code.x`,-`Indicator Code.y`)
```

The structure of the Population data (using str())shows that all
variables except Population is Character, Population variable is in
numeric. here the as.factor function is used to convert the Country
code, Indicator Name and Country Name into factor variables by checking
the unique values using the unique() function. The level() functions
shows the different levels for the updated factor variables. The
indicator name variable is further labelled as Male and Female for
"Population,male and"Population,female" and ordered. the class()
indicates an ordered variable which shows that the conversion is
successful.

The same steps are followed for the LE data as they have the same
variables.

The str() function shows the updated structure of both population and LE
and shows that the variables are successfully converted into factor
where necessary.

Once the structure of the data is corrected, the population and LE data
is merged to form the 'Merged' data using inner join function as inner
join help in identifying the common countries and variables between the
two data. In this case since the observations were same for both, there
were no deletion of data. Using the left or right join was also giving
the same result because of the common observations and variables.

To further tidy the data the indicator code column has been removed as
they are irrelevant to the study and most likely used as an internal
reference.

<br> <br>

## **Tidy & Manipulate Data II**

```{r}
#Creating a relative life expectancy variable

# calculating the mean global life expectancy
global_average_life_expectancy <- round(mean(Merged$`Life expectancy at birth`,na.rm= TRUE),2)

Merged$Relative_Life_Expectancy <- round(with(Merged, 
                                         `Life expectancy at birth` / global_average_life_expectancy),2)   


```

A new variable Relative Life expectancy is created here using the global
average life expectancy. the global average LE is calculated by the mean
of the LE across all countries at birth (na.rm - removes all NA in the
data). The relative frequency is the calculated by dividing the life
expectancy at birth for each country by the global average life
expectancy. This derived variable provides a measure of how each
country's life expectancy compares to the average. A value greater than
1 indicates a above average life expectancy for the country, a value
less than 1 indicates below average LE and equal to 1 indicates the LE
of the country is equal to the global average.

Now the merged dataset has 7 variables instead of the 6 including the
new Relative life expectancy variable

<br> <br>

## **Scan I**

```{r}
#scanning for missing values

colSums(is.na(Merged)) 

#There is missing values in Population , life expectancy at birth and relative life expectancy

# to see the na values in population
missing_population <- Merged %>% filter (., is.na(`Population`)== TRUE) 

#Since there are only 186 observation and is for the country 'not classified' and 'West bank and gaza', we can exclude these values. It also only accounts for 0.5% of the total data

Merged <- Merged[complete.cases(Merged$Population), ] 

colSums(is.na(Merged))

#missing values in life expectancy column
missing_LE <- Merged %>% filter (., is.na(Merged$`Life expectancy at birth`)== TRUE)

#Imputing values for LE column by using the mean
Merged$`Life expectancy at birth` <-impute(Merged$`Life expectancy at birth`, fun = mean)

colSums(is.na(Merged))
#Recalculating the Relative LE
global_average_life_expectancy <- round(mean(Merged$`Life expectancy at birth`,na.rm= TRUE),2)

Merged$Relative_Life_Expectancy <- round(with(Merged, 
                                         `Life expectancy at birth` / global_average_life_expectancy),2) 

colSums(is.na(Merged))

# Checking for infinite or NaN or NA values using a function called is.specialorNA

is.specialorNA <- function(x){
  if (is.numeric(x)) (is.infinite(x) | is.nan(x) | is.na(x))
}
sapply(Merged, function(x) sum(is.specialorNA(x) ))

#no infinite or NAN values

#checking for obvious inconsistencies
# Relative life expectancy has to be between 0 and 2
(Rule1 <- editset(c("`Relative_Life_Expectancy` >= 0", "`Relative_Life_Expectancy` <= 2")))

#Population of a country is expected to be more than 0
Violation1 <-violatedEdits(Rule1, Merged)
summary(Violation1)

(Rule2 <- editset(c("`Population` >= 0")))
Violation2 <-violatedEdits(Rule2, Merged)
summary(Violation2)
```

To identify if the data has any missing values colsums(is.na()) is used.
the results show that there are 186 missing values in population and
1782 values in LE and Relative LE. Upon close inspection of the missing
values in population by using a filter function, it can be seen that the
186 observations are in the country "not classified" and "West bank and
Gaza" and accounts for only 0.5% of the total data. there we can exclude
these values by using the complete.cases function.

The colsums(is.na()) function now shows that there are no more missing
values in Population. however there are still 1596 missing values in
Life expectancy at birth and Relative LE.To have a better understanding
of the missing value a variable called missing_LE is created which shows
the NAs are spread across countries. Since its not ideal to waste all
the information because of missing values, the best approach to dealing
with it will be to impute it with the mean value.

The colsums(is.na()) function now shows that there are no more missing
values in Life expectancy, however there are still 1596 missing values
in Relative LE. Since Relative LE is a derived variable, we can
recalculate the values using the formula used previously. The
colsums(is.na()) function now shows that there are no more missing
values

In order to check for infinite or NaN values a function called
is.specialorNA is created and then applied to the data using the sapply
function. The function shows that there are no NaN or special values in
the data.

In order to check for obvious inconsistencies, we have set up two rules
using the editset() function. One pertaining to the population stating
that the population of a country has to greater than 0 and the second
related to the relative life expectancy - stating that the Relative LE
should be between 0 and 2. The summary() function shows that there are
no violation detected for both the rules.

<br> <br>

## **Scan II**

```{r}
Merged$Population %>%  boxplot(main="Boxplot of Population", ylab="Population", col = "red")
Merged$`Life expectancy at birth` %>%  boxplot(main="Boxplot of Life expectancy", ylab="Population", col = "green")
Merged$Relative_Life_Expectancy%>%  boxplot(main="Boxplot of Relative Life expectancy", ylab="Population", col = "blue")

#checking for outliers using the Z-scores
z.scores <- Merged$Population %>%  scores(type = "z")
z.scores %>% summary()
length (which(abs(z.scores) >3 ))

z.scores <- Merged$`Life expectancy at birth` %>%  scores(type = "z")
z.scores %>% summary()
length (which(abs(z.scores) >3 ))

z.scores <- Merged$Relative_Life_Expectancy %>%  scores(type = "z")
z.scores %>% summary()
length (which(abs(z.scores) >3 ))

#scatter plot beterrn Life Expectancy and Population
Merged %>% plot( Population ~`Life expectancy at birth`,data = ., ylab="Population", xlab="LE", main="Population by Life expectancy")

# Imputing the outliers by capping
cap <- function(x) {
  quantiles <- quantile(x, c(0.05, 0.25, 0.75, 0.95))
  x[x < quantiles[2] - 1.5 * IQR(x)] <- quantiles[1]
  x[x > quantiles[3] + 1.5 * IQR(x)] <- quantiles[4]
  (x)
}

Subset <- Merged %>% dplyr::select(Population,`Life expectancy at birth`,Relative_Life_Expectancy)
head(Subset)

Merged_capped <- sapply(Subset, FUN = cap)
summary(Merged_capped)

```

One of the ways of detecting outlier is using the boxplot, in which
"Tukey's method of outlier detection" is used to detect outliers. The
box plot of population, life expectancy and Relative life expectancy
shows that they have outliers. In order to identify the number of
outliers z score method is used. According to this method if the
absolute value of its z-score is greater than 3, then it is regarded as
an outlier. From the summary of Population, we can see that the z-score
has a minimum of -0.3121 and maximum of 10.5034.The length() function
shows that the population variable has 776 outliers. Similarly, the Life
expectancy variable has 86 outliers and the Relative Life expectancy
variable has 81 outliers.

The scatter plot also shows that there are some possible outliers for
the pair of population and Life expectancy variables.

Since the number of outliers doesn't seem like they were due to any data
processing errors, deleting or imputing them with mean is not the right
strategy to handle it. In order to handle the outliers the capping or
Winsorising method is being used. A function called cap is created to
cap the values outside the limits. Since this function can be performed
only on numerical values, a subset is created for population, LE and
relative LE variables. Merged_capped data is created by applying the cap
function on the Subset data.The summary of the capped data shows the
updated summary statistics.

<br> <br>

## **Transform**

```{r}
# data transformation - to decrease the skewness and convert the distribution into a normal distribution
hist(Merged$Population)

#the log transformation
log10_population <- log10(Merged$Population)
hist(log10_population) # approximately normal

log_population <- log(Merged$Population)
hist(log_population) # log10 works better

sqrt_population <- sqrt(Merged$Population)
hist(sqrt_population) #still skewed

sqr_population <- Merged$Population^2 
hist(sqr_population)  #still skewed

recip_population <- 1/Merged$Population
hist(recip_population)  #still skewed

BoxCox_population<- BoxCox(Merged$Population,lambda = "auto")
hist(BoxCox_population)  #still skewed

#centring and scaling or z score standardisation
z_Population <- scale(Merged$Population, center = TRUE, scale = TRUE) 
hist(z_Population)

z_LE <- scale(Merged$`Life expectancy at birth`, center = TRUE, scale = TRUE) 
hist(z_LE)

z_relative_LE<- scale(Merged$Relative_Life_Expectancy, center = TRUE, scale = TRUE) 
hist(z_relative_LE)
```

The Transformation are done only the population variable since there is
a page limit on this project.The initial histogram of population shows
that its left skewed Different types of transformations are being done
to convert it to normal distribution such as log, lo10, sqrt,sqr, BoxCox
etc. It can be seen that the log10 and log transformations works best.
Log 10 is slightly better than the log transformations in this case.

The Z-score standardization is also done on all three variables to
preprocess the numerical variables. With centering, the distributions
are centered around a mean of 0 and with scaling has brought the scales
of variables between -4 and 2 for easy comparability <br> <br>

## **Link to Presentation**

The link to  the recording is given below:
https://rmit-arc.instructuremedia.com/embed/d481f6ea-e77e-4556-859a-0c0e4c8b2a4d

<br> <br>

## **References**
Sona Taheri (2023) Data Wrangling [Module 3 - 9 Demo and Notes], RMIT University,Melbourne
Studio. (n.d.). Retrieved from Canvas, accessed on 19 September 2023:https://rmit.instructure.com/accounts/1/external_tools/38?launch_type=global_navigation(https://rmit.instructure.com/accounts/1/external_tools/38?launch_type=global_navigation)

Bache S, Wickham H (2022).
magrittr: A Forward-Pipe Operator for R R package version 2.0.3,https://CRAN.R-project.org/package=magrittr(https://CRAN.R-project.org/package=magrittr)

Damian W. Betebenner (2021). randomNames: Function for Generating Random Names and a Dataset. (Rpackage version 1.5-0.0 URLhttps://cran.r-project.org/package=randomNames(https://cran.rproject.org/package=randomNames)

Harrell Jr F (2023).Hmisc: Harrell MiscellaneousR package version 5.1-1,https://R Core Team (2023)R: A Language and Environment for Statistical Computing R Foundation for StatisticalComputing, Vienna, Austria.
https://www.R-project.org/ (https://www.R-project.org/)

Wickham H, François R, Henry L, Müller K, Vaughan D (2023).
dplyr: A Grammar of Data Manipulation. Rpackage version 1.1.2,https://CRAN.R-project.org/package=dplyr(https://CRAN.Rproject.org/package=dplyr)

bank, W. (n.d.). Life expectancy at birth,male. Retrieved from data.worldbank: https://data.worldbank.org/indicator/SP.DYN.LE00.MA.IN?locations=AU

bank, W. (n.d.). Population,male. Retrieved from data.worldbank: https://data.worldbank.org/indicator/SP.POP.TOTL.MA.IN?locations=AU

bank, W. (n.d.). Life expectancy at birth,female. Retrieved from data.worldbank:https://data.worldbank.org/indicator/SP.DYN.LE00.FE.IN?locations=AU

bank, W. (n.d.). Population,female. Retrieved from data.worldbank: https://data.worldbank.org/indicator/SP.POP.TOTL.FE.IN?locations=AU

