---
title: "Generating and Preprocessing synthetic data for Woolworths" 
author: "Deepthi Suresh"
date: "17 September 2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default 
editor_options: 
  markdown: 
    wrap: 72
---

## **Setup**

```{r}
library(dplyr)
library(magrittr)
library(randomNames)
library(Hmisc)

```
## **Video presentation**
Link : https://rmit-arc.instructuremedia.com/embed/a7bcd5c8-c9c9-4468-8746-c9b5a08c120b

## **Data generation**

```{r}
#Creating data set 1 called Inventory Data set
#Variable 1 - Product ID
Product_id <- c(1:100)
head(Product_id)

#Variable 2 - Product Category
Category <- c( 	"Fruits and vegetables", 
                "bakery",
                "poultry", 
                "dairy eggs & fridge",
                "Pantry",
                "Freezer",
                "Beauty",
                "Home & Lifestyle",
                "Cleaning and Maintenance",
                "Pet")
set.seed(3455)
Product_Category <- sample(Category,100,replace = TRUE)
head(Product_Category)

#creating a draft of the Inventory dataset
inventory_dataset <- data.frame(Product_id,Category)

# variable 3 - Product Names
#creating vectors with different product names for each Category
FV<- c("onions","tomato", "apples","banana","garlic" )
B <- c("cookies","cupcakes","croissant","bread","cakes")
P <- c("poultry","meat","seafood","BBQ meat")
D <- c("cheese","milk","eggs","dips","yoghurt")
PT <- c( "tea & coffee","baking","Herbs","oil","pasta")
Fr <- c("Ice cream", "frozen fruits","frozen meals","frozen meat","frozen pizza")
Be <- c("sun protection","dental care", "hair care","skin care")
HL <- c("Dining","party Supplies","Storage","Home Decor","Toys")
CM<- c("laundry","kitchen","pest control","hardware")
pet <- c("cat & kitten","Dog & puppy","Bird, Fish & Small & pets")

set.seed(799)
inventory_dataset <- inventory_dataset %>%
  mutate(Product_name = ifelse(Category=="Fruits and vegetables",sample(FV,nrow(inventory_dataset),replace = TRUE), 
            ifelse(Category == "bakery",sample(B,nrow(inventory_dataset),replace= TRUE),
            ifelse(Category == "poultry",sample(P,nrow(inventory_dataset),replace= TRUE),
            ifelse(Category == "Pantry",sample(PT,nrow(inventory_dataset),replace= TRUE), 
          ifelse(Category == "Freezer",sample(Fr,nrow(inventory_dataset),replace= TRUE),
          ifelse(Category == "Beauty",sample(Be,nrow(inventory_dataset),replace= TRUE),
          ifelse(Category == "dairy eggs & fridge",sample(D,nrow(inventory_dataset),replace= TRUE),
          ifelse(Category == "Home & Lifestyle",sample(HL,nrow(inventory_dataset),replace= TRUE),
                                                                             ifelse(Category == "Cleaning and Maintenance",sample(CM,nrow(inventory_dataset),replace= TRUE),
                                                                                    ifelse(Category == "Pet",sample(pet,nrow(inventory_dataset),replace= TRUE),0)))))))))))
                                          
head(inventory_dataset$Product_name)                                          
#Variable no 4 -Quantity in stock
set.seed(98786)
Quantity_stock <- round(runif(100,min = 30, max = 150),0) 

head(Quantity_stock)         
# variable no 5 - supplier
set.seed(67588)
Supplier <- randomNames(100)
head(Supplier)
# Combining the variables to form the inventory data set
inventory_dataset <- data.frame(inventory_dataset,Quantity_stock,Supplier)
head(inventory_dataset)

#Adding missing values to the Supplier Variable by introducing a rand variable
set.seed(5647)
inventory_dataset %<>% 
  mutate(rand = runif(100, min = 0, max = 1))
inventory_dataset %<>% 
  mutate(Supplier = case_when(rand >= 0.05 ~ Supplier))

# removing rand column
inventory_dataset<- inventory_dataset[,-6]

sum(is.na(inventory_dataset$Supplier))

head(inventory_dataset)

#Generating Dataset 2 called Sales Dataset 
#Variable 1: Order No
set.seed(321434)
Order_No <- sample(round(runif(50, min = 40000, max = 50000),0),100,replace = TRUE)
head(Order_No)
#Variable 2 : Sale date
set.seed(4456)
date <- Sys.Date() - sort(sample(1:1000, 100))
head(date)
#Variable 3: Quantity
set.seed(234656)
Sales_Quantity <- round(runif(100,min = 1, max = 25),0)
head(Sales_Quantity)

# Variable 4: Price per unit, after discount
set.seed(79789)
Price <- round(runif(100,min = 1, max = 25),2)
head(Price)
#Variable 5: Payment Methods
set.seed(475684)
Payment_method <- sample(1:4,100,replace = TRUE)
head(Payment_method)
#Variable 6: Discount Amount 
set.seed(73245)
Discount <- sample(round(runif(50, min = 0, max = 15),2),100,replace = TRUE)
head(Discount)
#Variable 7: Productid
productid <- sample(round(runif(100, min = 0, max = 100),0),100,replace = FALSE)
head(productid)
#generating a temporary dataset
Sales_Dataset <- data.frame(Order_No,date,productid, Sales_Quantity,Price,Discount,Payment_method)

#variable 8: Total sales
# assuming pricing errors/importing errors and estimating 5 as sd
Sales_Dataset <- Sales_Dataset %>%
  mutate(Total_Sales = (Price * Sales_Quantity))
# rounding the sales data to 2 decimals
Sales_Dataset$Total_Sales <- round(Sales_Dataset$Total_Sales,2)
#Generating a plot to check the correlation
plot(x = Sales_Dataset$Total_Sales, 
     y = Sales_Dataset$Sales_Quantity,
     main = "Correlation between Sales and Quantity", 
     xlab = "Total Sales", 
     ylab = "Quantity Sold")

#Adding Missing values in discount by using a rand variable
set.seed(9865)
Sales_Dataset %<>% 
  mutate(rand = runif(100, min = 0, max = 1))
Sales_Dataset %<>% 
  mutate(Discount = case_when(rand >= 0.05 ~ Discount))
# removing rand column
Sales_Dataset<- Sales_Dataset[,-9]
sum(is.na(Sales_Dataset$Discount)) # no of missing values

head(Sales_Dataset)
```

I have considered two datsets from Woolworths ( the data is only a
subset of the original dataset because of its sheer size.) The first
dataset is the Inventory dataset containing information regarding the
various product categories and product names that Woolworths carry,
along with their supplier name and Stock in Hand.the datset consists of
100 observations and 5 variables as given below:

Product_id(NUM) - unique identifier of the products available

Category(CHR) - broad categorisation of the product

Product_Name(CHR) - individual product names

Quantity_stock(NUM) - the stock in hand for each product

Supplier(CHR) - the name of the supplier providing the products to
Woolworths

In order to generate the variables, the following steps were taken:

Product_Id was generated using a vector of 1-100. A random sample from a
vector containing a subset of Categories in Woolworths were introduced
to generate the product category variable using the 'sample' function.
In order to create the product names under each category, 10 vectors
were created, from which a random sample would be assigned to each
observation under each category using the 'Ifelse' and 'sample'
function. Quantity \_stock was generated using the 'runif' function
keeping an assumption that stores have lead times and require a minimum
of 30 units to order a new batch and maximum of 100 ( scaling it down
for the purpose of this assignment). 'Suppliers' variable was created
using the 'randomnames' function. Once the variables were created it was
combined to form a dataframe called Inventory_Dataset using the
'data.frame' function.

In order to induce missing values, the supplier variable was chosen. A
random variable 'rand' was first created using the runif function.then a
'case_when' function was applied on supplier variable with a condition
on the random variable of "\>= 0.05". This then deletes 5% of the
suppliers from the dataset,selected at random by converting to NAs where
the 'rand' value is less than 0.05.

The second dataset is a Sales dataset of Woolworths. It consists of 100
observations with 8 variables as given below:

Order_No(NUM) - the transaction number for the purchase

Sale_Date(DATE) - the date on which the sale happened

Sales_Quantity(NUM) - the no of units purchased

Price(NUM) - the price of the products after discounts

Payment Methods(INT) - the method of payment used by the customers like
Cash,card, online or giftcards

Discount(NUM) - the discount offered by Woolworths on individual items

Product_id(NUM) - id of the items purchased

Total_Sales(NUM) - correlated variable calculated using a formula
indicating the total sales generated by the store

In order to generate the variables the following steps have been taken:

Order_No is generated using random sampling from a vector containing
numbers between 40000 and 50000 ( assuming the order numbers are a 5
digit code at Woolworths and we are creating only a sample of 100
observationa). Sales_Date is generated keeping the system date( using
the 'sys.date' and random dates using the' runif' and 'sample' function.
Quantity and Price are generated in a similar way using the 'runif' and
'sample' function. An assumption of maximum 25 quantities can be
purchased, is maintained and the price range is assumed to be 1-25 AUD
for the products in the dataset. The payment methods available to the
customers are cash,credit card, online payments and gift cards-
generated using the sample function. The discount that is offered is
assumed to be capped at 15\$ per product and a random sample is
generated. Productid is a random sample of observations from 1-100.

In order to generate the correlated variable, the following formula is
used

Total_Sales = (Price \* Sales_Quantity).

An assumption of 5 SD is maintained for this variable. A scatter plot
for the variables Quantity sold and Total Price are generated which
shows a mildly positive correlation, which could mean that more the
quantity sold more would be the total sales that are generated.

All the variables are then combined to form the Sales_Dataset

In order to add missing values in the dataset, the discount variable has
been selected as this is not a variable that is being used for
calculations in this particular study. Missing value is introduced using
a random variable called 'rand' and 'case_when' function with a criteria
of 'rand \>= 0.05'. This then deletes 5% of the Discount from the
dataset,selected at random by converting to NAs where the 'rand' value
is less than 0.05.'Sum(is.na()' function is used to calculate the no of
NA values in the discount variable

The 'head' of sales and Inventory dataset provides a glimpse of the
newly created dataset

## **Merging data sets** 

```{r}

merged1 <- left_join(inventory_dataset,Sales_Dataset,by = join_by
                       (Product_id==productid))
head(merged1,10)


```

The sales and inventory dataset is then joined using the left join
function with the key variable of Product_id in Inventory Dataset and
productid from the sales Dataset. Since we have used a left join all the
variable from the Inventory dataset will be a part of the merged data
set but only the common variables from the Sales data will be part of
the new data set. The new merged1 datset has 137 observations with 12
variables.

## **Checking structure of combined data** 

```{r}

# str of the dataset
head(merged1)
dim(merged1)
str(merged1)                     
names(merged1)

#converting to factor
# converting payment methods from INT to Factor
unique(Payment_method)
merged1$Payment_method <- merged1$Payment_method %>%
  factor(.,levels = c( 1,2,3,4),
         labels = c("credit card","gift card","online payment","cash"))
class(merged1$Payment_method)
levels(merged1$Payment_method)
# converting Category from CHR to Factor
unique(Category)
merged1$Category <- merged1$Category%>%
  factor(.,levels = c("Fruits and vegetables","bakery","poultry" ,                
                    "dairy eggs & fridge","Pantry","Freezer",                
                      "Beauty","Home & Lifestyle","Cleaning and Maintenance",
                      "Pet" ))
class(merged1$Category)
levels(merged1$Category)
# converting product_name from CHR to Factor
unique(merged1$Product_name)
merged1$Product_name <- merged1$Product_name%>%
  factor(.,levels = c( "apples","bread","seafood","cheese","Herbs","frozen pizza",             
                       "sun protection","party Supplies","kitchen","Bird, Fish & Small & pets","garlic","croissant",               
                       "meat","baking","frozen meat","skin care","Dining","cat & kitten",             
                       "banana", "cookies","milk","oil","frozen meals","Home Decor",               
                       "hardware","Dog & puppy","onions","BBQ meat","dips","tea & coffee",             
                       "dental care","Storage","cakes","yoghurt","Toys","poultry",                  
                       "pasta","Ice cream","hair care","pest control","tomato","frozen fruits"))

class(merged1$Product_name)
levels(merged1$Product_name)

str(merged1)

```

The structure of the merged dataset shows that there are 137
observations and 12 variables. The category, product name and payment
methods seems like ideal candidates to convert into Factor variable. The
'names' function shows the names of the different variables

In order to convert the payment method to factor variable, we have first
identified the unique values in the variable using the 'unique' function
and then we have used the 'factor' function along with levels to
identify the different levels 1,2,3,4 and the corresponding labels for
each of these levels such as "credit card", "gift card", "online
payment", "cash". Once the factor conversion is done, the 'class'
function confirms that the payment method is converted into factor and
the 'levels' function shows the new labels. The above process is
followed for the category variable and product names.

The 'str' function shows the updated structure of the merged dataset
with category, product name and payment methods being a Factor variable.

## **Generate summary statistics**

```{r}
# summarising the total sales based on category
Summary_TS <-merged1 %>% 
  group_by(Category) %>%
  summarise(min = round(min(Total_Sales,na.rm = TRUE),2),
                   first_quantile = round(quantile(Total_Sales, 0.25,na.rm = TRUE),2),
                   Median = round(median(Total_Sales,na.rm = TRUE),2),
                   Mean = round(mean(Total_Sales,na.rm = TRUE),2),
                   third_quantile = round(quantile(Total_Sales,0.75,na.rm = TRUE),2),
                   max = round(max(Total_Sales,na.rm = TRUE),2), 
                   stand_dev = round(sd(Total_Sales,na.rm = TRUE),2), 
                   n = n(),
                   missing = sum(is.na(Total_Sales)))%>% 
  ungroup()
head(Summary_TS,10)

# summarising the total sales based on Payment methods
Summary_PM <-merged1 %>% 
  group_by(Payment_method) %>% 
  summarise(min = round(min(Total_Sales,na.rm = TRUE),2),
            first_quantile = round(quantile(Total_Sales, 0.25,na.rm = TRUE),2),
            Median = round(median(Total_Sales,na.rm = TRUE),2),
            Mean = round(mean(Total_Sales,na.rm = TRUE),2),
            third_quantile = round(quantile(Total_Sales,0.75,na.rm = TRUE),2),
            max = round(max(Total_Sales,na.rm = TRUE),2), 
            stand_dev = round(sd(Total_Sales,na.rm = TRUE),2), 
            n = n(),
            missing = sum(is.na(Total_Sales)))%>% na.omit(Summary_PM) %>%
  ungroup()

head(Summary_PM)

```

In order to summarize the data, we have performed two summary statistics
on Total sales based on category and Payment Methods

The summarize function is used here to generate summary statistics and
saved into 'Summary_TS' data. The total sales is calculated based on the
10 category that we have used in the dataset. The following summary
statistics are being used here - min, Quartile 1, Median, Mean, Quartile
3, Max, Standard Deviation, N and no of missing values. We have also
used Na.rm function to remove the NA values for numerical calculations.
The same process have been followed to create 'Summary_PM', a summary
statistics for Total sales grouped by payment method.

On an average majority of the payment are made through online payment,
followed by giftcard, credit card and cash.The maximum amount
spent is using cash worth of 550$ whereas the least amount spent
is 3.64$.

According to the Summary_TS data, highest average sale is from bakery
category followed by Freezer.Pantry and poultry seems to be a low
performing category with a mean of 129.30 and 132.83.Max sale in a
single purchase is from the Beauty and Dairy department worth of 548$
and 550$

## **Scanning data** 

```{r}
# Using head to check the variables with missing values
head(merged1)
colSums(is.na(merged1))

#Imputing missing values of Discount using mean vale
merged1$Discount <- round(impute(merged1$Discount,fun = mean),2)
head(merged1)
# bringing back Na in discount where there are no order details
merged1$Discount[is.na(merged1$Order_No)]<-NA 

# deleting supplier values with missing values  
merged1 <-merged1[complete.cases(merged1),]

# checking the dataset
head(merged1)

str(merged1)

```

The head function is used here to take a look at at the merged data.
'ColSums' function is used to count the total no of missing values in
each column ( Is.na function is not used here because of the number of
observations) The missing values in the Discount variable is dealt by
imputing NAs with the mean of discount using the 'impute'function.
However when we do this we notice that the values are filled for all
observations including the one with no order number. In order to remove
this, we use the 'is.na' function on the order_no variable.

The missing values in Supplier variable is removed by deleting them as
they account for only \<5% of the data and will not impact any of the
calculations in this study. we have used the 'complete.cases' function
to do this.

The str function shows that after removing 5% of missing values there are now 95 observations with 12 variables

## **References** 

R Core Team (2023). *R: A Language and Environment for Statistical
Computing*. R Foundation for Statistical Computing, Vienna, Austria.
<https://www.R-project.org/>.

Wickham H, François R, Henry L, Müller K, Vaughan D (2023). *dplyr: A
Grammar of Data Manipulation*. R package version 1.1.2,
<https://CRAN.R-project.org/package=dplyr>.

Bache S, Wickham H (2022). *magrittr: A Forward-Pipe Operator for R*. R
package version 2.0.3, <https://CRAN.R-project.org/package=magrittr>

Damian W. Betebenner (2021). randomNames: Function for Generating Random
Names and a Dataset. (R package version 1.5-0.0 URL
<https://cran.r-project.org/package=randomNames>

Harrell Jr F (2023). *Hmisc: Harrell Miscellaneous*. R package version
5.1-1, <https://CRAN.R-project.org/package=Hmisc>

Group, W. (2023, sept 18). Woolworths. Retrieved from Discover -
Shopping Online, Accessed on 18 September 2023,
<https://www.woolworths.com.au/shop/discover/shopping-online/pickup?utm_source=google&utm_medium=cpc&utm_campaign=WW-0001&cq_net=g&cq_src=GOOGLE&cq_cmp=Woolies_8458_BAU_Brand_Fulfill_WW-0001&cq_med=71700000097600800&cq_plac=&cq_term=woolworths%20online&ds_>

Sona Taheri (2023) Data Wrangling [Module 4, 5 Demo and Notes &
"Generating Sataset Module], RMIT University,Melbourne

Studio. (n.d.). Retrieved from Canvas, accessed on 19 September 2023: https://rmit.instructure.com/accounts/1/external_tools/38?launch_type=global_navigation
